---
title: "Pumping Test Analysis Results"
output: 
    html_document:
         number_sections: true
params:
  current.ptest: NA
---

```{r load_lib,echo=FALSE}
# The `params` object is available in the document.
library(pumpingtest)
library(knitr)
library(kableExtra)
```

# Basic information about the Pumping Test

The pumping test  `r params$current.ptest$id[[1]]` dataset contains `r length(params$current.ptest$t)` measurements and a short summary of their statistical properties is included in the following table:
```{r data_summary,echo=FALSE}
t <- params$current.ptest$t
s <- params$current.ptest$s
ptest.df <- data.frame("Time" = t, "Drawdown" = s)
summary(ptest.df)
```


# Diagnostic Plot

```{r plot_diagnostic, echo=FALSE, fig.width=6,fig.height=5,fig.align='center'}
plot(params$current.ptest, type = "diagnostic", dmethod = "smoothspline")
```

# Aquifer Parameter Estimation

```{r plot_estimation, echo=FALSE, fig.width=6,fig.height=5,fig.align='center'}
print(params$current.ptest$estimated)
print(params$current.ptest$hydraulic_parameters)
plot(params$current.ptest, type = "estimation", dmethod = "smoothspline", 
     results = FALSE)
```


# Model Diagnostic

The quality of the estimated model can be checked using the following figure:

```{r plot_model_diagnostic, echo=FALSE, fig.width=6,fig.height=5,fig.align='center'}
plot(params$current.ptest, type = "model.diagnostic")
```


# Appendix{-}

```{r data_table, echo=FALSE}
kable(ptest.df, format = "html", digits = 3, align = c('c','c'), caption = "VES DATA") %>%   kable_styling(bootstrap_options = "striped", full_width = F)
```
